\pkg{dataMaid}: your assistant for documenting supervised data quality screening in \proglang{R}

### 2018-04-11 latex/article_accept.tex
Formatting corrections for JSS

%Virker ikke:
%\newcommand{\R}{\proglang{R}\xspace}

\newcommand{\R}[1]{\code{#1}}

@@

While \pkg{dataMaid}s primary raison d'être is auto-generating data
quality assessment overview documents, we still wish to emphasize that
it is \emph{not} a tool for unsupervised data cleaning. This qualifies
as the second paradigm of \pkg{dataMaid}: Data cleaning decisions
should always be made by humans. Therefore, \pkg{dataMaid} does not
supply any tools for ``fixing'' errors in the data. However, we do
provide interactive functions that can be used to identify potentially
erroneous entries in a dataset and that can make it easier to solve
data issues, one variable at a time.

A number of \proglang{R} packages made for other pre-analysis steps
are already available, including \pkg{janitor} \citep{janitor},
\pkg{assertive} \citep{assertive}, \pkg{dplyr} \citep{dplyr},
\pkg{tidyr}\citep{tidyr}, \pkg{data.table} \citep{data.table},
\pkg{DataCombine} \citep{DataCombine}, \pkg{validate}
\citep{validate}, and \pkg{assertr} \citep{assertr}. These packages
focus on different stages of the pre-analysis work.  \pkg{janitor}
provides tools for data import with a particular emphasis on the
challenges of getting neat data frames from Microsoft Excel data
files. \pkg{dplyr}, \pkg{tidyr}, \pkg{data.table} and
\pkg{DataCombine} go a few steps further by providing a wide array of
extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) and \pkg{assertive} offer tools for identifying errors
in a dataset by checking the state of the variable given a set of
pre-specified rules, and their focus is on internal validity rather
than general data screening. In practice, this means that quite elegant
tools for, e.g., linear restraints among the variables in a dataset
can be applied, but looking for potentially miscoded missing values is
not really feasible. The main difference between these two challenges
is the direction in which the data is inspected: While linear
constraints work observation-wise with no ambiguity, determining
whether or not something is a miscoded missing value often requires
knowledge about the full variable (e.g. range or data type), and thus
it should be performed variable-wise. \pkg{validate} does not
currently allow for user-defined extensions of the latter type,
thereby limiting its data cleaning potential.  Automatic data
correction functions are also provided by \pkg{validate} which we
consider to be quite a dangerous cocktail: all power is given to the
the computer with no human supervision, and investigators are less
likely to make an active, case-specific choice regarding the handling
of the potential errors.  Finally, no tools have been made to easily
document exactly which checks and preliminary results were used in the
data cleaning process. The \pkg{assertr} package provide very similar
--- and very nice --- tools to those of \pkg{validate}, but without
any amibitions of conducting auto-cleaning.

@@

One last package that should be mentioned in this context is
\pkg{DataExplorer} \citep{DataExplorer}. While this package does not
address data cleaning issues \emph{per se}, its general strategy is
quite similar to that of \pkg{dataMaid} and to the paradigms presented
above. This package provides a few simple, but practical, tools for
exploratory data analysis, including automated
documentation. Therefore, we find \pkg{DataExplorer} to be a good
candidate for a next-step package after data cleaning is finished.

@@

\begin{Schunk}
\begin{Sinput}
R> library("dataMaid")
R> data("toyData")
R> toyData
\end{Sinput}
\begin{Soutput}
# A tibble: 15 x 6
     pill events region     change     id spotifysong
   <fctr>  <dbl> <fctr>      <dbl> <fctr>      <fctr>
 1    red      1      a -0.6264538      1  Irrelevant
 2    red      1      a  0.1836433      2  Irrelevant
 3    red      1      a -0.8356286      3  Irrelevant
 4    red      2      a  1.5952808      4  Irrelevant
 5    red      2      a  0.3295078      5  Irrelevant
 6    red      6      b -0.8204684      6  Irrelevant
 7    red      6      b  0.4874291      7  Irrelevant
 8    red      6      b  0.7383247      8  Irrelevant
 9    red    999      c  0.5757814      9  Irrelevant
10    red     NA      c -0.3053884     10  Irrelevant
11   blue      4      c  1.5117812     11  Irrelevant
12   blue     82      .  0.3898432     12  Irrelevant
13   blue     NA        -0.6212406     13  Irrelevant
14   <NA>    NaN  other -2.2146999     14  Irrelevant
15   <NA>      5  OTHER  1.1249309     15  Irrelevant
\end{Soutput}
\end{Schunk}

@@

By default, an \proglang{R} markdown file and a rendered pdf,  word or html
overview document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three
parts, two of which are presented in Figure~\ref{fig:example1}. First,
an overview of what was done is presented under the title \textit{Data
  report overview}. Secondly, an index listing each variable along
with an indication of whether it was found to be problematic or not is
provided. Thirdly, as seen in Figure~\ref{fig:example2}, each variable
in the dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution when relevant and a list of flagged
issues, if any. For instance, as shown in Figure~\ref{fig:example2},
for the \code{numeric}-type variable \code{events} from
\code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and
\code{NaN}), while two values were also flagged as potential outliers
that should be investigated more carefully.

The arguments to \code{makeDataReport()} can be used to modify the
contents and the look of the data report according to the user's
needs. The most commonly used arguments are summarized in
Table~\ref{table.cleanFormals} and they are grouped according to the
part of the data assesment and report generation they influence. In
order to understand this distinction, a glimpse of the inner structure
of \code{makeDataReport()} is shown in
Figure~\ref{figure:cleanStructure}. Below, we present a few examples
on how to use the arguments from Table \ref{table.cleanFormals} to
influence the output of a \code{makeDataReport()} call.

@@

\smallskip Control input variables, looks and meta information\\
\quad \code{useVar} & What variables should be used? & \code{NULL} (corresponding to all variables) \\
\quad \code{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \code{"asIs"} \\
\quad \code{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \code{FALSE} \\
%\quad \code{listChecks} & Should an overview of what checks were performed be listed in the \textit{Data report overview}? &  \code{TRUE} \\
%slettet for at få plads til nye argumenter
\quad \code{preChecks} & What check functions should be called to determine whether a variable is suitable for summarization, visualization and checking? & \code{c("isKey", "isSingular", "isSupported")} \\
\quad \code{reportTitle} & What should the title displayed on the front page of the report be? & \code{NULL} (corresponds to the dataset name) \\
\quad \code{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \code{TRUE} \\

\smallskip Control summarize, visualize, and check steps \\
\quad \code{summaries} & What summaries should be performed for each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{visuals} & What type of visualization should be provided for each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{checks} & What checks should be applied to each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{mode} & What steps should be performed for each variable (out
                 of the three possibilities \textit{summarize},
                 \textit{visualize}, \textit{check})? &
                                                        \code{c("summarize", "visualize", "check")}  \\
\quad \code{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \code{TRUE} \\
\quad \code{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \code{10} \\
\quad \code{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \code{2} \\
\quad \code{treatXasY} & How should non-supported variable classes be handled? & \code{NULL} (no handling) \\
%Takes a list on the format \code{list(X = Y)} where \code{X} is a non-supported variable class (e.g. \code{complex}) and \code{Y} is a supported one (e.g. \code{numeric}) & \code{NULL} (No rules for handling non-supported variable types) \\

@@

 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & \blue{$\times$} & & \blue{$\times$} & \blue{$\times$} \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed whitespace &  \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isSingular} & Check if the variable contains only a single value & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$}  \\
 \quad \code{isKey} & Check if the variable is a key & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \smallskip   \\
 \quad \code{isSupported} & Check if the variable is among the supported variable types & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \smallskip   \\
 \hline
\end{tabular}
\caption{Overview of all summary-, visual- and  check functions currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \code{character}, \code{factor},
  \code{integer}, \code{labelled}, \code{logical} (boolean), \code{numeric}, and \code{Date}, respectively. The default settings of \code{makeDataReport()} are marked in blue.}
\label{table.SVCfunctions}
\end{table}

@@

As with \code{summaryFunction}s, a complete list of available
\code{checkFunction}s is obtained by calling
\code{allCheckFunctions()}. Note however, that \code{checkFunction}s have a
usage beyond the \code{checks} arguments, namely in the
\textit{pre-check} stage. In this stage, it is determined whether or
not each variable is suitable for the summarize/visualize/check (SVC)
steps. The functions used in the pre-check stage should be
\code{checkFunction}s that are applicable to all variable classes. The
default pre-checks, the functions \code{isKey()}, \code{isSingular()} and \code{isSupported()}, check
whether a variable has unique values for all observations, only a
single value for all observations, and is not among the variable types supported by \pkg{dataMaid}, respectively. If one of these
statements are true, the variable will not be subjected to the SVC
steps.  We can allow singular variables to move on to the SVC step by
only checking for keys and non-supported variables in the pre-check step:

@@

In summary, and as indicated in Figure~\ref{figure:cleanStructure}, there are two stages
where \code{makeDataReport()} applies functions to each of the variables:
\begin{enumerate}
\item In the pre-check stage.
\item As part of the summarize/visualize/check (SVC) steps.
\end{enumerate}
Each of these stages are controllable using appropriate function
arguments in \code{makeDataReport()}, and above we have shown examples
of how to tweak them to modify the data cleaning outputs. However, if
the dataset at hand requires new, additional checks, then more control
is needed. The package contains a vignette that explains the details of how to modify and
expand the possibilities by producing new summary, visual, and check
functions.

One might also encounter datasets with variables that are not among the 7 classes mentioned in the above (\code{character}, \code{Date}, \code{factor}, \code{integer}, \code{labelled}, \code{logical} and \code{numeric}), for instance variables of type \code{complex} or user-defined classes. It is possible to tell \code{makeDataReport()} how to handle such variables by use of the argument \code{treatXasY}. This argument takes a list where the names correspond to "new" variable types (\code{X}), while the entries must be supported variable types (\code{Y}). For instance, we can instruct \code{dataMaid} to treat complex variables as numerics and generate a data report for a type \code{complex} variable like this:

@@

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=5]{dataMaid_presidentData_appB.pdf}}
\frame{\includegraphics[width=7.5cm,page=6]{dataMaid_presidentData_appB.pdf}}
\end{center}
\caption{Part of the data overview report generated for the \code{presidentData} dataset. Here, we see pages 4 and 5 from the \textit{Variable list}. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix2}.}
\label{fig:bigExampleP45}
\end{figure}

We will now put the bits and pieces from above together and show how \code{makeDataReport()} can be used on a less artificial dataset to create a useful overview report and how the interactive tools can subsequently be used to assist the actual data cleaning process. More specifically, we will create a report describing the \code{presidentData} dataset, which is available in \pkg{dataMaid} and use the information from this report to clean up the data. \code{presidentData} is a sligtly mutilated dataset with information about the 45 first US presidents, but with a few common data issues and a blind passenger. The dataset contains one observation per president and has the following variables:
\begin{description}
\item[\code{lastName}] The last name of the president.
\item[\code{firstName}] The first name of the president.
\item[\code{orderOfPresidency}] The number in the order of presidents.
\item[\code{birthday}] The birthday of the president.
\item[\code{stateOfBirth}] The state in which the president was born.
\item[\code{assassinationAttempt}] Was there an assassination attempt on the president?
\item[\code{sex}] The sex of the president.
\item[\code{ethnicity}] The ethnicity of the president.
\item[\code{presidencyYears}] The duration of the presidency.
\item[\code{ageAtInauguration}] The age of the president at inauguration.
\item[\code{favoriteNumber}] The favourite number of the president (fictional).
\end{description}

\begin{Schunk}
\begin{Sinput}
R> data(presidentData)
R> head(presidentData)
\end{Sinput}
\begin{Soutput}
    lastName firstName orderOfPresidency   birthday  stateOfBirth
1 Washington    George                 1 1732-02-22      Virginia
2      Adams      John                 2 1735-10-30 Massachusetts
3  Jefferson    Thomas                 3 1743-04-13      Virginia
4    Madison     James                 4 1751-03-16      Virginia
5     Monroe     James                 5 1758-04-28      Virginia
6      Adams      John                 6 1767-07-11 Massachusetts
  assassinationAttempt  sex ethnicity presidencyYears
1                    0 Male Caucasian               7
2                    0 Male Caucasian               3
3                    0 Male Caucasian               8
4                    0 Male Caucasian               8
5                    0 Male Caucasian               8
6                    0 Male Caucasian               4
  ageAtInauguration favoriteNumber
1                57    3+0.000000i
2                61    4+0.000000i
3                57    0+1.414214i
4                57   10+0.000000i
5                58    3+0.000000i
6                57    9+0.000000i
\end{Soutput}
\end{Schunk}

We discuss the results of a data overview report generated for this dataset below, but first there are a few special features of the dataset and wishes for the data report that require us to customize it using some of the arguments for \code{makeDataReport}. We have the following points of interest:
\begin{enumerate}
\item A couple of variables are used to store names, namely \code{lastName} and \code{firstName}. In order to use special bibliographical analysis tools on these, and only these, variables, it might be convenient to assign them a special class. Therefore, these variables have been set to have class \code{Name} by use of the base \proglang{R} function \code{class()}. When we wish to make a data report for the dataset, we have to tell \code{makeDataReport()} how to handle such \code{Name}-type variables, as they are not among the supported variable types mentioned in the above. This can be done using the \code{treatXasY} argument.
\item We use the \code{character} class for a few categorical variables that have a lot of different levels, but where it is not a data mistake that these levels each generally have very few observations, e.g. the variable \code{stateOfBirth}. Therefore, we would like to disable the \code{identifyLoners} check (which flags variables with <6 observations in any of the levels) for this variable type. This can be done using the \code{checks} argument.
\item We would like the report to be called "Dirty president data" in order to reflect that it is, indeed, a report concerning dirty data about presidents.
\end{enumerate}
Incorporating these three customizations, we can load the data and generate a report by calling:

@@

The first four pages of the resulting report can be inspected in Figures \ref{fig:bigExampleP01} and \ref{fig:bigExampleP23}. Note that all the customized settings can be identified from the first two pages, without having to read through the report: The new title is written on the front page, the check settings are displayed in the \textit{Checks performed} table and the strategy for handling \code{Name} variables is documented below this table. The full data report, except for the front page, is available in Appendix \ref{sec:appendix2}.

@@

On page 3, we see the contents for the three first variables. Here, we identifify a prefixed whitespace in the lastname entry for President Truman and we find that a dot was entered as a first name; this is a typical choice for coding missing values in e.g. Stata, and therefore, it is flagged as a potential miscoded missing value. The variable \code{orderOfPresidency} is not summarized, visualized or checked because it is categorical and contains unique values for each observation.

>>>>

%Virker ikke:
%\newcommand{\R}{\proglang{R}\xspace}
% \newcommand{\R}[1]{\code{#1}}

@@

While \pkg{dataMaid}s primary raison d'être is auto-generating data
quality assessment overview documents, we still wish to emphasize that
it is not a tool for unsupervised data cleaning. This qualifies
as the second paradigm of \pkg{dataMaid}: Data cleaning decisions
should always be made by humans. Therefore, \pkg{dataMaid} does not
supply any tools for ``fixing'' errors in the data. However, we do
provide interactive functions that can be used to identify potentially
erroneous entries in a dataset and that can make it easier to solve
data issues, one variable at a time.

A number of \proglang{R} packages made for other pre-analysis steps
are already available, including \pkg{janitor} \citep{janitor},
\pkg{assertive} \citep{assertive}, \pkg{dplyr} \citep{dplyr},
\pkg{tidyr}\citep{tidyr}, \pkg{data.table} \citep{data.table},
\pkg{DataCombine} \citep{DataCombine}, \pkg{validate}
\citep{validate}, and \pkg{assertr} \citep{assertr}. These packages
focus on different stages of the pre-analysis work.  \pkg{janitor}
provides tools for data import with a particular emphasis on the
challenges of getting neat data frames from Microsoft Excel data
files. \pkg{dplyr}, \pkg{tidyr}, \pkg{data.table} and
\pkg{DataCombine} go a few steps further by providing a wide array of
extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) and \pkg{assertive} offer tools for identifying errors
in a dataset by checking the state of the variable given a set of
pre-specified rules, and their focus is on internal validity rather
than general data screening. In practice, this means that quite elegant
tools for, e.g., linear restraints among the variables in a dataset
can be applied, but looking for potentially miscoded missing values is
not really feasible. The main difference between these two challenges
is the direction in which the data is inspected: While linear
constraints work observation-wise with no ambiguity, determining
whether or not something is a miscoded missing value often requires
knowledge about the full variable (e.g., range or data type), and thus
it should be performed variable-wise. \pkg{validate} does not
currently allow for user-defined extensions of the latter type,
thereby limiting its data cleaning potential.  Automatic data
correction functions are also provided by \pkg{validate} which we
consider to be quite a dangerous cocktail: all power is given to the
the computer with no human supervision, and investigators are less
likely to make an active, case-specific choice regarding the handling
of the potential errors.  Finally, no tools have been made to easily
document exactly which checks and preliminary results were used in the
data cleaning process. The \pkg{assertr} package provide very similar
--- and very nice --- tools to those of \pkg{validate}, but without
any ambitions of conducting auto-cleaning.

@@

One last package that should be mentioned in this context is
\pkg{DataExplorer} \citep{DataExplorer}. While this package does not
address data cleaning issues per se, its general strategy is
quite similar to that of \pkg{dataMaid} and to the paradigms presented
above. This package provides a few simple, but practical, tools for
exploratory data analysis, including automated
documentation. Therefore, we find \pkg{DataExplorer} to be a good
candidate for a next-step package after data cleaning is finished.

@@

\begin{Schunk}
\begin{Sinput}
R> library("dataMaid")
R> data("toyData")
R> toyData
\end{Sinput}
\begin{Soutput}
# A tibble: 15 x 6
   pill  events region change id    spotifysong
   <fct>  <dbl> <fct>   <dbl> <fct> <fct>      
 1 red     1.00 a      -0.626 1     Irrelevant 
 2 red     1.00 a       0.184 2     Irrelevant 
 3 red     1.00 a      -0.836 3     Irrelevant 
 4 red     2.00 a       1.60  4     Irrelevant 
 5 red     2.00 a       0.330 5     Irrelevant 
 6 red     6.00 b      -0.820 6     Irrelevant 
 7 red     6.00 b       0.487 7     Irrelevant 
 8 red     6.00 b       0.738 8     Irrelevant 
 9 red   999    c       0.576 9     Irrelevant 
10 red    NA    c      -0.305 10    Irrelevant 
11 blue    4.00 c       1.51  11    Irrelevant 
12 blue   82.0  .       0.390 12    Irrelevant 
13 blue   NA    " "    -0.621 13    Irrelevant 
14 <NA>  NaN    other  -2.21  14    Irrelevant 
15 <NA>    5.00 OTHER   1.12  15    Irrelevant 
\end{Soutput}
\end{Schunk}

@@

By default, an \proglang{R} markdown file and a rendered pdf,  word or html
overview document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three
parts, two of which are presented in Figure~\ref{fig:example1}. First,
an overview of what was done is presented under the title ``Data
  report overview''. Secondly, an index listing each variable along
with an indication of whether it was found to be problematic or not is
provided. Thirdly, as seen in Figure~\ref{fig:example2}, each variable
in the dataset is presented in turn using (up to) three tools in the
``Variable list'': A table summarizing key features of the
variable, a figure visualizing its distribution when relevant and a list of flagged
issues, if any. For instance, as shown in Figure~\ref{fig:example2},
for the \code{numeric}-type variable \code{events} from
\code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and
\code{NaN}), while two values were also flagged as potential outliers
that should be investigated more carefully.

The arguments to \code{makeDataReport()} can be used to modify the
contents and the look of the data report according to the user's
needs. The most commonly used arguments are summarized in
Table~\ref{table.cleanFormals} and they are grouped according to the
part of the data assessment and report generation they influence. In
order to understand this distinction, a glimpse of the inner structure
of \code{makeDataReport()} is shown in
Figure~\ref{figure:cleanStructure}. Below, we present a few examples
on how to use the arguments from Table \ref{table.cleanFormals} to
influence the output of a \code{makeDataReport()} call.

@@

\smallskip Control input variables, looks and meta information\\
\quad \code{useVar} & What variables should be used? & \code{NULL} (corresponding to all variables) \\
\quad \code{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \code{"asIs"} \\
\quad \code{onlyProblematic} & Should only variables flagged as problematic be included in the ``Variable list''? & \code{FALSE} \\
%\quad \code{listChecks} & Should an overview of what checks were performed be listed in the "Data report overview"x? &  \code{TRUE} \\
%slettet for at få plads til nye argumenter
\quad \code{preChecks} & What check functions should be called to determine whether a variable is suitable for summarization, visualization and checking? & \code{c("isKey", "isSingular", "isSupported")} \\
\quad \code{reportTitle} & What should the title displayed on the front page of the report be? & \code{NULL} (corresponds to the dataset name) \\
\quad \code{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \code{TRUE} \\

\smallskip Control summarize, visualize, and check steps \\
\quad \code{summaries} & What summaries should be performed for each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{visuals} & What type of visualization should be provided for each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{checks} & What checks should be applied to each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{mode} & What steps should be performed for each variable (out
                 of the three possibilities ``summarize'',
                 ``visualize'', ``check'')? &
                                                        \code{c("summarize", "visualize", "check")}  \\
\quad \code{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \code{TRUE} \\
\quad \code{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \code{10} \\
\quad \code{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \code{2} \\
\quad \code{treatXasY} & How should non-supported variable classes be handled? & \code{NULL} (no handling) \\
%Takes a list on the format \code{list(X = Y)} where \code{X} is a non-supported variable class (e.g. \code{complex}) and \code{Y} is a supported one (e.g. \code{numeric}) & \code{NULL} (No rules for handling non-supported variable types) \\

@@

 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & \blue{$\times$} & & \blue{$\times$} & \blue{$\times$} \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed white space &  \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{isCPR} & Identify Danish personal identification numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isSingular} & Check if the variable contains only a single value & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$}  \\
 \quad \code{isKey} & Check if the variable is a key & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \smallskip   \\
 \quad \code{isSupported} & Check if the variable is among the supported variable types & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \smallskip   \\
 \hline
\end{tabular}
\caption{Overview of all summary-, visual- and  check functions currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \code{character}, \code{factor},
  \code{integer}, \code{labelled}, \code{logical} (boolean), \code{numeric}, and \code{Date}, respectively. The default settings of \code{makeDataReport()} are marked in blue.}
\label{table.SVCfunctions}
\end{table}

@@

As with \code{summaryFunction}s, a complete list of available
\code{checkFunction}s is obtained by calling
\code{allCheckFunctions()}. Note however, that \code{checkFunction}s have a
usage beyond the \code{checks} arguments, namely in the
pre-check stage. In this stage, it is determined whether or
not each variable is suitable for the summarize/visualize/check
steps. The functions used in the pre-check stage should be
\code{checkFunction}s that are applicable to all variable classes. The
default pre-checks, the functions \code{isKey()}, \code{isSingular()} and \code{isSupported()}, check
whether a variable has unique values for all observations, only a
single value for all observations, and is not among the variable types supported by \pkg{dataMaid}, respectively. If one of these
statements are true, the variable will not be subjected to the SVC
steps.  We can allow singular variables to move on to the SVC step by
only checking for keys and non-supported variables in the pre-check step:

@@

In summary, and as indicated in Figure~\ref{figure:cleanStructure}, there are two stages
where \code{makeDataReport()} applies functions to each of the variables:
\begin{enumerate}
\item In the pre-check stage.
\item As part of the summarize/visualize/check steps.
\end{enumerate}
Each stage is controllable using appropriate function
arguments in \code{makeDataReport()}, and above we have shown examples
of how to tweak them to modify the data cleaning outputs. However, if
the dataset at hand requires new, additional checks, then more control
is needed. The package contains a vignette that explains the details of how to modify and
expand the possibilities by producing new summary, visual, and check
functions.

One might also encounter datasets with variables that are not among the 7 classes mentioned in the above (\code{character}, \code{Date}, \code{factor}, \code{integer}, \code{labelled}, \code{logical} and \code{numeric}), for instance variables of type \code{complex} or user-defined classes. \code{makeDataReport()} can be told how to handle such variables by use of the argument \code{treatXasY}. This argument takes a list where the names correspond to "new" variable types (\code{X}), while the entries must be supported variable types (\code{Y}). For instance, we can instruct \code{dataMaid} to treat complex variables as numerics and generate a data report for a type \code{complex} variable like this:

@@

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=5]{dataMaid_presidentData_appB.pdf}}
\frame{\includegraphics[width=7.5cm,page=6]{dataMaid_presidentData_appB.pdf}}
\end{center}
\caption{Part of the data overview report generated for the \code{presidentData} dataset. Here, we see pages 4 and 5 from the ``Variable list''. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix2}.}
\label{fig:bigExampleP45}
\end{figure}

We will now put the bits and pieces from above together and show how \code{makeDataReport()} can be used on a less artificial dataset to create a useful overview report and how the interactive tools can subsequently be used to assist the actual data cleaning process. More specifically, we will create a report describing the \code{presidentData} dataset, which is available in \pkg{dataMaid} and use the information from this report to clean up the data. \code{presidentData} is a slightly mutilated dataset with information about the 45 first US presidents, but with a few common data issues and a blind passenger. The dataset contains one observation per president and has the following variables:
\begin{description}
\item[\code{lastName}] The last name of the president.
\item[\code{firstName}] The first name of the president.
\item[\code{orderOfPresidency}] The number in the order of presidents.
\item[\code{birthday}] The birthday of the president.
\item[\code{stateOfBirth}] The state in which the president was born.
\item[\code{assassinationAttempt}] Was there an assassination attempt on the president?
\item[\code{sex}] The sex of the president.
\item[\code{ethnicity}] The ethnicity of the president.
\item[\code{presidencyYears}] The duration of the presidency.
\item[\code{ageAtInauguration}] The age of the president at inauguration.
\item[\code{favoriteNumber}] The favorite number of the president (fictional).
\end{description}

\begin{Schunk}
\begin{Sinput}
R> data("presidentData")
R> head(presidentData)
\end{Sinput}
\begin{Soutput}
    lastName firstName orderOfPresidency   birthday  stateOfBirth
1 Washington    George                 1 1732-02-22      Virginia
2      Adams      John                 2 1735-10-30 Massachusetts
3  Jefferson    Thomas                 3 1743-04-13      Virginia
4    Madison     James                 4 1751-03-16      Virginia
5     Monroe     James                 5 1758-04-28      Virginia
6      Adams      John                 6 1767-07-11 Massachusetts
  assassinationAttempt  sex ethnicity presidencyYears
1                    0 Male Caucasian               7
2                    0 Male Caucasian               3
3                    0 Male Caucasian               8
4                    0 Male Caucasian               8
5                    0 Male Caucasian               8
6                    0 Male Caucasian               4
  ageAtInauguration favoriteNumber
1                57    3+0.000000i
2                61    4+0.000000i
3                57    0+1.414214i
4                57   10+0.000000i
5                58    3+0.000000i
6                57    9+0.000000i
\end{Soutput}
\end{Schunk}

We discuss the results of a data overview report generated for this dataset below, but first there are a few special features of the dataset and wishes for the data report that require us to customize it using some of the arguments for \code{makeDataReport}. We have the following points of interest:
\begin{enumerate}
\item A couple of variables are used to store names, namely \code{lastName} and \code{firstName}. In order to use special bibliographical analysis tools on these, and only these, variables, it might be convenient to assign them a special class. Therefore, these variables have been set to have class \code{Name} by use of the base \proglang{R} function \code{class()}. When we wish to make a data report for the dataset, we have to tell \code{makeDataReport()} how to handle such \code{Name}-type variables, as they are not among the supported variable types mentioned in the above. This can be done using the \code{treatXasY} argument.
\item We use the \code{character} class for a few categorical variables that have a lot of different levels, but where it is not a data mistake that these levels each generally have very few observations, e.g., the variable \code{stateOfBirth}. Therefore, we would like to disable the \code{identifyLoners} check (which flags variables with <6 observations in any of the levels) for this variable type. This can be done using the \code{checks} argument.
\item We would like the report to be called "Dirty president data" in order to reflect that it is, indeed, a report concerning dirty data about presidents.
\end{enumerate}
Incorporating these three customizations, we can load the data and generate a report by calling:

@@

The first four pages of the resulting report can be inspected in Figures \ref{fig:bigExampleP01} and \ref{fig:bigExampleP23}. Note that all the customized settings can be identified from the first two pages, without having to read through the report: The new title is written on the front page, the check settings are displayed in the ``Checks performed'' table and the strategy for handling \code{Name} variables is documented below this table. The full data report, except for the front page, is available in Appendix \ref{sec:appendix2}.

@@

On page 3, we see the contents for the three first variables. Here, we identify a prefixed white space in the last name entry for President Truman and we find that a dot was entered as a first name; this is a typical choice for coding missing values in e.g., Stata, and therefore, it is flagged as a potential miscoded missing value. The variable \code{orderOfPresidency} is not summarized, visualized or checked because it is categorical and contains unique values for each observation.
